---
title: "Code for mapping CSO habitat suitability"
author: "Liraz Bistritz"
date: "2025-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(plyr)
library(dplyr)
library(tidyr)
library(tidyverse)
library(terra)
library(raster)
library(sf)
```

###California Spotted Owl (CSO) habitat:###

##In order for the code to work, the following needs to be true:
#The size raster's values should be:
size 1's value is 10
size 2's values is 20
size 3's values is 30
size 4's values is 40
size 5's values is 50

#In the canopy cover, the values should be 1-4, where:
D is 4
M	is 3
X	is 0
P	is 2
S is 1

#The veg types need to be numerical too, in this current version it was as follows:
DFR is 100
EPN is 200
JPN	is 300
LPN	is 400
MHC	is 500
MHW	is 600
MRI	is 700
PPN	is 800
RFR	is 900
SMC	is 1000
WFR is 1100

#make sure these values align with the values in the base layer you use, i.e. that DFR is coded as 100 etc.
#then the 3 rasters above need to be summed, so that for example the value 154 in this summed raster means: veg type 100, size 50, canopy cover 4.
#once you have this summed base layer, you can run the code:

```{r}
dir <- "~"  #set the directory where the boundary, species' ranges and base layer are located
output_dir <- "~" #set the directory to which your outputs will be saved

#the boundary:
bound <- rast(paste0(dir, "/ACCEL_boundary_TealeAlbers.tif")) #the landscape boundary raster

#the base layer (the summed raster):
base_layer <- rast(paste0(dir, "/base_layer.tif")) %>%
  terra::project(bound, method="near") %>%
  terra::crop(bound)
plot(base_layer)

#CSO range, to clip the habitat layers by it (can be downloaded from https://map.dfg.ca.gov/metadata/ds0897.html)
cso_range <- rast(paste0(dir, "/b270.tif")) %>%
  terra::project(bound, method="near") %>%
  terra::crop(bound)
```

```{r}
#creating a binary layer for habitat that is highly suitable for nesting:
cso_habitat_1 <- c(154, 254, 354, 454, 554, 654, 754, 854, 954, 1054, 1154) #these are: "DFR5D", "EPN5D", "JPN5D", "LPN5D", "MHC5D", "MHW5D", "MRI5D", "PPN5D", "RFR5D", "SMC5D", "WFR5D
ones <- c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1)
n_cso_unique_hab_ID <- data.frame(cso_habitat_1, ones) #creating a data frame
n_cso_unique_hab_ID

n_cso_ras <- classify(base_layer, n_cso_unique_hab_ID, others=0) #reclassifying the suitable values into 1, to create the binary habitat layer
plot(n_cso_ras)
n_cso_ras <- ifel(cso_range == 1, n_cso_ras, 0)  #keeping only suitable habitat within the CSO range
plot(n_cso_ras)
n_cso_ras <- mask(n_cso_ras, bound)
plot(n_cso_ras)

writeRaster(n_cso_ras, paste0(output_dir, "/FVEG_CSO_suitable_habitat_1.tif"), overwrite=T, datatype='INT1S') #saving the output


##################
#removing outputs from R environment before running the next lines:
rm(n_cso_unique_hab_ID, n_cso_unique_hab_ID2, cso_habitat_1, cso_habitat_2, cso_habitat_4, cso_habitat_5, cso_filt_final_5, f_cso_unique_hab_ID, f_cso_unique_hab_ID2)

#creating a binary layer for habitat that is moderately suitable for nesting:
cso_habitat_2 <- c(153, 253, 353, 453, 553, 653, 753, 853, 953, 1053, 1153) #these are: "DFR5M", "EPN5M", "JPN5M", "LPN5M", "MHC5M", "MHW5M", "MRI5M", "PPN5M", "RFR5M", "SMC5M", "WFR5M"
ones <- c(1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1)
n_cso_unique_hab_ID2 <- data.frame(cso_habitat_2, ones) #creating a data frame

n_cso_ras2 <- classify(base_layer, n_cso_unique_hab_ID2, others=0) #reclassifying the suitable values into 1, to create the binary habitat layer

n_cso_ras2 <- ifel(cso_range == 1, n_cso_ras2, 0)  #keeping only suitable habitat within the CSO range

n_cso_ras2 <- mask(n_cso_ras2, bound)
plot(n_cso_ras2)
writeRaster(n_cso_ras2, paste0(output_dir, "/FVEG_CSO_suitable_habitat_2.tif"), overwrite=T, datatype='INT1S') #saving the output


##################
#removing outputs from R environment before running the next lines:
rm(n_cso_unique_hab_ID, n_cso_unique_hab_ID2, cso_habitat_1, cso_habitat_2, cso_habitat_4, cso_habitat_5, cso_filt_final_5, f_cso_unique_hab_ID, f_cso_unique_hab_ID2)

#creating a binary layer for habitat that is highly suitable for foraging:
cso_habitat_4 <- c(144, 544, 644, 744, 844, 944, 1044, 1144) #"DFR4D", "MHC4D", "MHW4D", "MRI4D", "PPN4D", "RFR4D", "SMC4D", "WFR4D"
ones <- c(1, 1, 1, 1, 1, 1, 1, 1)
f_cso_unique_hab_ID <- data.frame(cso_habitat_4, ones) #creating a data frame

f_cso_ras <- classify(base_layer, f_cso_unique_hab_ID, others=0) #reclassifying the suitable values into 1, to create the binary habitat layer

f_cso_ras <- ifel(cso_range == 1, f_cso_ras, 0) #keeping only suitable habitat within the CSO range
plot(f_cso_ras)
f_cso_ras <- mask(f_cso_ras, bound)
plot(bound)

writeRaster(f_cso_ras, paste0(output_dir, "/FVEG_CSO_suitable_habitat_4.tif"), overwrite=T, datatype='INT1S') #saving the output


##################
#removing outputs from R environment before running the next lines:
rm(n_cso_unique_hab_ID, n_cso_unique_hab_ID2, cso_habitat_1, cso_habitat_2, cso_habitat_4, cso_habitat_5, cso_filt_final_5, f_cso_unique_hab_ID, f_cso_unique_hab_ID2)

#creating a binary layer for habitat that is moderately suitable for foraging:
cso_habitat_5 <- c(143, 543, 643, 743, 843, 943, 1043, 1143) #these are: "DFR4M", "MHC4M", "MHW4M", "MRI4M", "PPN4M", "RFR4M", "SMC4M", "WFR4M"
ones <- c(1, 1, 1, 1, 1, 1, 1, 1)
f_cso_unique_hab_ID2 <- data.frame(cso_habitat_5, ones) #creating a data frame

f_cso_ras2 <- classify(base_layer, f_cso_unique_hab_ID2, others=0) #reclassifying the suitable values into 1, to create the binary habitat layer

f_cso_ras2 <- ifel(cso_range == 1, f_cso_ras2, 0) #keeping only suitable habitat within the CSO range
plot(f_cso_ras2)
f_cso_ras2 <- mask(f_cso_ras2, bound)
plot(bound)

writeRaster(f_cso_ras2, paste0(output_dir, "/FVEG_CSO_suitable_habitat_5.tif"), overwrite=T, datatype='INT1S') #saving the output

```

```{r}
#bringing the four habitat layers together:

habitat <- n_cso_ras + n_cso_ras2 + f_cso_ras + f_cso_ras2 #sum layer values
habitat_final <- ifel(habitat > 0, 1, 0) #if the summed layer is greater than 0, classify the pixel as 1, otherwise, 0.

plot(n_cso_ras)
plot(n_cso_ras2)
plot(f_cso_ras)
plot(f_cso_ras2)
plot(habitat_final)

writeRaster(habitat_final, paste0(output_dir, "/FVEG_CSO_suitable_habitat_1245.tif"), overwrite=T, datatype='INT1S') #saving the output
```
